Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-09 paper/paper.tex
figures are almost final
Updated conclusions, Future work sections and fixed some issues discussed with Amey

The strongly regular grammar is converted into a set of
NFAs,   one  for   each  $\var{\Lanv{i}{x}}$.   The  $\hookrightarrow$
simplification  is now  done  on the  NFAs  by repeatedly  introducing
$\epsilon$  edges  to  bypass  pairs  of  consecutive  edges  labelled
\bcar\acar\ or \bcdr\acdr\  and constructing the
$\epsilon$-closure. Since the $\epsilon$-closure step may create more
adjacent \bcar\acar\ and \bcdr\acdr\  edges, the simplification is
 continued  till a  fixed point is  reached. The algorithm  for this
appears  in~\cite{asati14lgc},   Figure[].  Finally  we   replace  the
$\clazy$  edges  by  $\epsilon$  and determinize  the  automaton  thus
formed.

@@

Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the  grammar for  \var{\Df{\length}{1'}}.  The  NFA is
simplified by replacing the edges labeled \clazy\ by $\epsilon$. While
checking for liveness during GC, a forward access-path traversing such
an edge  will be accepted  only if it  can reach a final  state.  This
effectively implements  the $\hookrightarrow$ simplification  rule for
\clazy for forward paths. The resulting automaton, made deterministic,
is shown in Figure~\ref{fig:example-automata}(b). This expectedly says
that  given  a demand  $\epsilon$  on  \length,  the liveness  of  its
argument             is             $\acdr^{*}$.            Similarly,
Figure~\ref{fig:example-automata}(c)    shows   the    automaton   for
\var{\Lanv{9}{\pa}}. The simplified automaton, shown in
Figure~\ref{fig:example-automata}(d) does not accept any forward
paths. This reflects the lazy nature of our language. \length\  does not
evaluate its elements and thus \pa\ is dead at $\pi_9$.

@@

     \Fn{\Copy{$\mathit{ref}$}}
        {
          \If{$\mathit{ref}$ is live}
             {
               $\mathit{newRef} = $  $\mathit{dupHeapCell(ref})$\;
               \If{cellType($\mathit{ref}$) is $cons$}
                  {
                    {
                      $\mathit{newCar}  = $  \Copy($\mathit{ref\!.car})$\;
                      $\mathit{newRef.car}   =  newCar$\;
                    }
                    {
                      $\mathit{newCdr}$ =   \Copy($\mathit{ref.cdr})$\;
                      $\mathit{newRef.cdr}   = \mathit{newCdr}$\;
                    }    
                  }
               \If {cellType($\mathit{ref}$) is closure}
                   {
                     $\mathit{newRef.arg1}$ =  \Copy$\mathit{(cell.arg1)}$\;
                     $\mathit{newRef.arg2}$ =  \Copy$\mathit{(cell.arg2)}$\;
                   }
             }
        }

@@

\subsection{Liveness-based collector with reachability-based closure
  collection}
\begin{algorithm}[t]
  \Fn{\Lgc{}}
     {
       \For {each reference $\mathit{ref}$ in root set}
            {copy($\mathit{ref}$)\;}
            $\mathit{copyReferencesOnPrintStack()}$\;  
     }
     \Fn{\LCopy{$\mathit{ref}$}}
        {
          \eIf{cellType$(\mathit{ref})$ is $cons$}
              {
                \eIf {$\mathit{ref}$ is live and has not been copied using
                  reachability}
                     {
                       $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
                       $\mathit{ newRef.car}   = \mathit{ newCar}$\;
                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.cdr)}$\;
                       $\mathit{ newRef.cdr}   = \mathit{newCar}$\;
                     }
                     { $\mathit{makenull(ref)}$\;}}
              {
                \If {cellType$\mathit{(ref)}$ is closure and 
                  has not been copied}
                    { 
                      $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
                    }
              }
        }
        \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
\end{algorithm}
  Since the first  algorithm has  significant bookkeeping
overheads,  we have  a  second GC  scheme  that not  use liveness  for
copying closures.  In this  scheme, the liveness automata is consulted
only for \CONS\ cells; we assume everything under a closure to be live
and copy everything  that is reachable. Algorithm~\ref{algo:lgc-b}  describes  this scheme.

@@

An advantage of  using reachablity for closures it  avoids the problem
of  multiple  explorations  from  the  same  cell.  This  is  done  by
maintaining  a flag in  each cell  to indicate  whether it  was copied
using  reachability.

@@

\subsection{Copying closure using liveness vs Copying closures using reachability}
To compare the effect of using reachability to copy closures instead of liveness we take a couple of programs from our benchmark suite and execute them with 
both. As shown in figure-add-figure-here, the number of garbage collections 
in both cases are same but the time taken varies largely. Both these observations can be explained as follows, since we approximate the liveness of closure to 
the liveness at its creation point, the liveness we get is usually grossly over approximated. 
Using this liveness may not give too much of a benefit compared to reachability as nearly everything would be live at its point of creation. 
As mentioned in section-ref, using liveness for closures may entail multiple traversals over the same structure consuming more time. 
Therefore we believe that using reachability for closures is a sweet spot in terms of performance for liveness based garbage collectors. 
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

@@

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by  our collector.  In case  of LGC  for eager
languages  the gap  was  very  narrow and  almost  touched the  actual
liveness curve.  In case of  lazy languages, due to using reachability
for copying we  end up copying more  cells. Any data which  is part of
closure irrespective  of whether it is  live or not gets  copied. Thus
many non-live  cells get  copied during garbage  collection increasing
the drag  time.  Our  experiments have  shown that  unless we  use the
exact  execution  point  for  closures bringing  down  the  number  of
non-live  cells  copied in  closures  cannot  be reduced.   Using  the
approximated liveness (liveness  at the creation point)  does not give
benefits  in terms  of non-live  cells inside  closures as  everything
would be live at the point of creation.
%% \comment{
%% The  increased effectiveness  of LGC  over RGC  is also  shown  in the
%% tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
%% provides  statistics  regarding the  analysis  itself.  The number  of
%% states and the analysis times  are within tolerable limits. Precision
%% of analysis refers to the percentage of dead cells that is collected by
%% LGC, averaged over all invocations.
%% The second table shows garbage  collection statistics for RGC and LGC.
%% LGC  collects larger garbage  per invocation,  drags cells  for lesser
%% time and requires a smaller heap size ({\em MinHeap}) for program to
%% run
%% in comparison with  RGC.

@@

\section{Related Work} Previous attempts to  increase  the space
efficiency
of functional programs by additional  reclamation of memory fall in two
broad categories. In the first,  the program itself is instrumented to
manage  reclamation and reallocation  without the  aid of  the garbage
collector.    Such    attempts   include:   sharing    analysis   based
reallocation~\cite{jones89compile},                       deforestation
techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
methods   that  enable   the   garbage  collector   to  collect   more
garbage~\cite{inoue88analysis,lee05static}   by
explicitly  nullifying  pointers  that  are not  live.   However,  the
nullification,  done   at  compile  time,   requires  sharing  (alias)
analysis.  Our  method, in contrast, does not  require sharing because
of the availability of the heap  itself at runtime.  To the best of our
knowledge, this is the first  attempt at liveness-based marking of the
heap during garbage collection.

@@

\section{Conclusions}
\label{sec:conclusion}
We  have  defined  a  notion  of liveness  on  structured  data;  this
generalizes classical liveness and strong liveness.  We started with a
general fully context-sensitive analysis  which we proved correct with
respect to  a minefield semantics  (this models the effect  of garbage
collection between every evaluation step).

To  avoid scalability  issues (and  to  avoid performing  part of  the
liveness computation at run time)  we defined an 0-CFA version of this
liveness analysis  in which  demands for function  $f$ at  all calling
contexts are conflated into  a single demand $\sigma_f$.  This enabled
us to treat the liveness equations symbolically obtaining context-free
grammars for liveness at each GC point (calls to user functions and to
$\CONS$).  These were then converted to DFAs for run-time consultation
by  the garbage collector.  Experiments confirm  the precision  of the
analysis.

To obtain performance figures we compared a reachability-based garbage
collector with a liveness-based  collector.  This showed a decrease in
the  number  of  GCs,   more  garbage  collected  per  invocation.   A
significant benefit of LGC is  that programs can run in smaller memory
when compared to  RGC. This is potentially useful  in situations where
memory  is  limited---as with  embedded  systems.  For  a majority  of
programs, the garbage collection times were reduced.

One issue we  highlighted was that while fewer  nodes were marked (and
more garbage collected), sometimes  cons cells could be visited and
traversed  multiple times  with different  sets of  liveness  paths to
explore.  Further work includes  improvements to the classical copying
collector to reduce the cost of this.

@@

%Mention lower peak memory usage.
%Mention faster GC than full LGC and more garbage collected than RGC.
%Pitch it as sweet spot between full LGC and RGC.
%Future work
%Mention full LGC
%Mention k-liveness, use length example to show 1-liveness is not
sufficient.
%Mention mixed mode GC,
% - doing RGC most of the times and switching to LGC in case RGC fails
to collect any garbage
% - doing LGC most of the times but regularly checking the reachability
of the heap and
% switching to RGC if LGC almost equals RGC

>>>>

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification  is  continued  till  a fixed  point  is  reached.  The
algorithm for this appears in~\cite{asati14lgc}, Figure[].  Finally we
replace the  $\clazy$ edges by  $\epsilon$ and check if  the automaton
has  valid path  (consisting  of only  \acar\ and  \acdr)  to a  final
state. If it has a valid path,  we remove the \clazy\ edge and mark the
source node as  final, else we remove the \clazy\  edge without marking
the source  node as final.   Once the  \clazy\ edges are  eliminated we
convert the automaton into a deterministic automaton.

@@

Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the  grammar for  \var{\Df{\length}{1'}}.  The  NFA is
simplified by replacing the edges labeled \clazy\ by $\epsilon$. While
checking for liveness during GC, a forward access-path traversing such
an edge  will be accepted  only if it  can reach a final  state.  This
effectively implements  the $\hookrightarrow$ simplification  rule for
\clazy\ for forward paths. The resulting automaton, made deterministic,
is shown in Figure~\ref{fig:example-automata}(b). This expectedly says
that  given  a demand  $\epsilon$  on  \length,  the liveness  of  its
argument             is             $\acdr^{*}$.            Similarly,
Figure~\ref{fig:example-automata}(c)    shows   the    automaton   for
\var{\Lanv{9}{\pa}}. The simplified automaton, shown in
Figure~\ref{fig:example-automata}(d) does not accept any forward
paths. This reflects the lazy nature of our language. \length\  does not
evaluate its elements and thus \pa\ is dead at $\pi_9$.

@@

     \Fn{\Copy{$\mathit{ref}$}}
        {
          \If{$\mathit{ref}$ is found live using dfa}
             {
               $\mathit{newRef} = $  $\mathit{dupHeapCell(ref})$\;
               \If{cellType($\mathit{ref}$) is $cons$}
                  {
                    {
                      $\mathit{newCar}  = $  \Copy($\mathit{ref\!.car})$\;
                      $\mathit{newRef.car}   =  newCar$\;
                    }
                    {
                      $\mathit{newCdr}$ =   \Copy($\mathit{ref.cdr})$\;
                      $\mathit{newRef.cdr}   = \mathit{newCdr}$\;
                    }    
                  }
               \If {cellType($\mathit{ref}$) is closure}
                   {
                     $\mathit{newRef.arg1}$ =  \Copy$\mathit{(cell.arg1)}$\;
                     $\mathit{newRef.arg2}$ =  \Copy$\mathit{(cell.arg2)}$\;
                   }
             }
        }

@@

\subsection{Liveness-based collector with reachability-based closure
  collection}
\begin{algorithm}[t]
  \Fn{\Lgc{}}
     {
       \For {each reference $\mathit{ref}$ in root set}
            {\LCopy($\mathit{ref}$)\;}
            $\mathit{copyReferencesOnPrintStack()}$\;  
     }
     \Fn{\LCopy{$\mathit{ref}$}}
        {
          \eIf{cellType$(\mathit{ref})$ is $cons$}
              {
                \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
                  reachability}
                     {
                       $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
                       $\mathit{ newRef.car}   = \mathit{ newCar}$\;
                       $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
                       $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
                     }
                     { $\mathit{makenull(ref)}$\;}}
              {
                \If {cellType$\mathit{(ref)}$ is closure and 
                  has not been copied}
                    { 
                      $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
                    }
              }
        }
        \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
\end{algorithm}
  Since the first  algorithm has  significant bookkeeping
overheads,  we have  a  second GC  scheme  that not  use liveness  for
copying closures.  In this  scheme, the liveness automata is consulted
only for \CONS\ cells; we assume everything under a closure to be live
and copy everything  that is reachable. Algorithm~\ref{algo:lgc-b}  describes  this scheme.

@@

Using  reachablity  for  closures   avoids  the  problem  of  multiple
explorations from the  same cell.  This is done by  maintaining a flag
in each cell to indicate whether it was copied using reachability.

@@

\subsection{Copying closure using liveness vs Copying closures using reachability}
To compare the  effect of using reachability to  copy closures instead
of liveness we take a couple  of programs from our benchmark suite and
execute them with both. As shown in \cred{figure-add-figure-here}, the
number of  garbage collections  in both  cases are  same but  the time
taken  varies largely.  Both these  observations can  be explained  as
follows, since we approximate the  liveness of closure to the liveness
at its  creation point, the  liveness we  get is usually  grossly over
approximated.  Using this liveness may not  give too much of a benefit
compared to  reachability as  nearly everything would  be live  at its
point of  creation.  As mentioned  in section-ref, using  liveness for
closures  may  entail  multiple  traversals over  the  same  structure
consuming more time.  Therefore we believe that using reachability for
closures is  a sweet spot in  terms of performance for  liveness based
garbage collectors.
%%
%% \comment{
%% The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
%% in the heap over time for RGC and LGC---here time is measured in terms
%% of the number of cons cells allocated. In addition, they also show the
%% number of  reachable cells and the  number of cells  that are actually
%% live\footnote{This   is  statically   approximated  by   our  liveness
%%   analysis}.  Since  the programs have  different memory requirements,
%% we have tuned the size of heap for each program to ensure a reasonable
%% number of  collections. An invocation  of RGC decreases the  number of
%% cells  in heap  till  it touches  the  curve of  reachable cells.   An
%% invocation of LGC decreases the number  of heap cells to no lower than
%% the curve of live cells.

@@

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by  our collector.  In case  of LGC  for eager
languages  the gap  was  very  narrow and  almost  touched the  actual
liveness curve.  In case of  lazy languages, due to using reachability
for copying we  end up copying more  cells. Any data which  is part of
closure irrespective  of whether it is  live or not gets  copied. Thus
many non-live  cells get  copied during garbage  collection increasing
the drag  time.  Our  experiments have  shown that  unless we  use the
exact  execution  point  for  closures, bringing  down  the  number  of
non-live  cells  copied in  closures  cannot  be reduced.   Using  the
approximated liveness (liveness  at the creation point)  does not give
benefits  in terms  of non-live  cells inside  closures as  everything
would be live at the point of creation.
%% \comment{
%% The  increased effectiveness  of LGC  over RGC  is also  shown  in the
%% tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
%% provides  statistics  regarding the  analysis  itself.  The number  of
%% states and the analysis times  are within tolerable limits. Precision
%% of analysis refers to the percentage of dead cells that is collected by
%% LGC, averaged over all invocations.
%% The second table shows garbage  collection statistics for RGC and LGC.
%% LGC  collects larger garbage  per invocation,  drags cells  for lesser
%% time and requires a smaller heap size ({\em MinHeap}) for program to
%% run
%% in comparison with  RGC.

@@

\section{Related Work} 
\cred{Need to rewrite this entire section need to probably mention some of these 
papers at least i found by googling liveness based GC,\\
http://www.cs.hmc.edu/~oneill/papers/Simplifiers-MSPC.pdf\\
http://costa.ls.fi.upm.es/papers/costa/AlbertGGZ12.pdf\\
http://hirzels.com/martin/papers/ecoop01-gc-liveness.pdf\\
}
Previous attempts to  increase  the space
efficiency
of functional programs by additional  reclamation of memory fall in two
broad categories. In the first,  the program itself is instrumented to
manage  reclamation and reallocation  without the  aid of  the garbage
collector.    Such    attempts   include:   sharing    analysis   based
reallocation~\cite{jones89compile},                       deforestation
techniques~\cite{wadler88deforest,gill93ashort,chitil99deforest},
methods  based  on   linear  logic~\cite{hofmann00linear}  and  region
analysis~\cite{tofte98region}.   Closer  to  our approach,  there  are
methods   that  enable   the   garbage  collector   to  collect   more
garbage~\cite{inoue88analysis,lee05static}   by
explicitly  nullifying  pointers  that  are not  live.   However,  the
nullification,  done   at  compile  time,   requires  sharing  (alias)
analysis.  Our  method, in contrast, does not  require sharing because
of the availability of the heap  itself at runtime.  To the best of our
knowledge, this is the first  attempt at liveness-based marking of the
heap during garbage collection.

@@

\section{Conclusions}
\label{sec:conclusion}
%% We  have  defined  a  notion  of liveness  on  structured  data;  this
%% generalizes classical liveness and strong liveness.  We started with a
%% general fully context-sensitive analysis  which we proved correct with
%% respect to  a minefield semantics  (this models the effect  of garbage
%% collection between every evaluation step).

%% To  avoid scalability  issues (and  to  avoid performing  part of  the
%% liveness computation at run time)  we defined an 0-CFA version of this
%% liveness analysis  in which  demands for function  $f$ at  all calling
%% contexts are conflated into  a single demand $\sigma_f$.  This enabled
%% us to treat the liveness equations symbolically obtaining context-free
%% grammars for liveness at each GC point (calls to user functions and to
%% $\CONS$).  These were then converted to DFAs for run-time consultation
%% by  the garbage collector.  Experiments confirm  the precision  of the
%% analysis.

%% To obtain performance figures we compared a reachability-based garbage
%% collector with a liveness-based  collector.  This showed a decrease in
%% the  number  of  GCs,   more  garbage  collected  per  invocation.   A
%% significant benefit of LGC is  that programs can run in smaller memory
%% when compared to  RGC. This is potentially useful  in situations where
%% memory  is  limited---as with  embedded  systems.  For  a majority  of
%% programs, the garbage collection times were reduced.

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
{\color{blue}In our current  work we have extended  the liveness based
  garbage collection to lazy languages  and show that it has benefits.
  Our  biggest   contribution  is  our  ability   to  handle  closures
  (suspended evaluations) during garbage collection.  We show that for
  a  practical  garbage collector  it  is  not possible  to  implement
  liveness based collector for  closures efficiently.  By implementing
  a mixed-mode garbage collection scheme, we were able to collect more
  cells than reachability based  collectors without paying the penalty
  of  a full  liveness based  collector.  We  believe this  is a  very
  practical idea and hence easily implementable.

We  also overcome  the drawback  of  multiple traversals  that a  full
liveness based  collector has  by using  an extra  bit and  setting it
whenever  the  cell   is  copied  using  reachability(as   part  of  a
closure). Our experimental results  have shown that such optimizations
provide performance benefits with very little loss of precision. Using
the  liveness  annotations  for  the  print  primitive  also  improves
performance and subsumes compile  time optimizations such as tail-call
optimization.

\subsection{Future work}
Although we  provide an  implementation for  a liveness  based garbage
collection scheme for lazy language, we  do not provide a formal proof
of its correctness.  A formal  proof of correctness would describe how
closures  are  handled  during  garbage  collection  and  liveness  is
propogated inside closures.

 Another interesting exercise is to narrow
the gap between the actual liveness  and the perceived liveness of our
garbage collector. Our  experiments show that a  significant number of
dead cells get trapped inside  closures. Using strictness analysis and
eagerly evaluating closures might release  more of these dead cells to
be  garbage  collected.   Since   our  garbage  collector  can  handle
closures(suspended evaluations)  another challenge would be  to extend
it to support higher order programs.

Orthogonally,  we can  improve the  efficiency of  the liveness  based
garbage  collector so  that  even  when the  decrease  in the  garbage
collections compared  to reachability is  not much, the  total garbage
collection is  not bad. Making  the DFA  smaller by using  better data
structures like Binary Decision Diagrams(BDD), truncating the liveness
automata  after   certain  depth(k-liveness)  are  some   ideas  worth
investigating.

 }

@@

%Mention lower peak memory usage.
%Mention faster GC than full LGC and more garbage collected than RGC.
%Pitch it as sweet spot between full LGC and RGC.
%Future work
%Mention full LGC
%Mention k-liveness, use length example to show 1-liveness is not
%sufficient.
%Mention mixed mode GC,
% - doing RGC most of the times and switching to LGC in case RGC fails
%to collect any garbage
% - doing LGC most of the times but regularly checking the reachability
%of the heap and
% switching to RGC if LGC almost equals RGC
